\vspace{1.0cm} \LARGE \bf Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex

### 2014-02-07 fluB.tex
Moving  supp to end of manuscript.
Also beginning to write some supplementary text.

\begin{figure}[h]
	\centering		
	\includegraphics[width=0.95\textwidth]{figures/RailroadPlotDated.pdf}
	\caption{\textbf{Schematic plot of reconstructed reassortments between Victoria and Yamagata lineage segments of influenza B virus.}
Lineages that coassort in genomes are represented by 8 parallel lines, with lineages that derive from the original Victoria clade colored red and lineages that derive from the original Yamagata clade colored blue.
Inter-lineage reassortment events are indicated by lines entering a different genome.
The angle of incoming lineages represents uncertainty in the timing of the event (mean date of the reassortant node and its parent node).
Lineage extinction dates are not shown accurately.
	\label{railroadPlot}
\end{figure}

@@

We have identified 3 instances of mixed-lineage PB1-PB2-HA reassortants from the data with the following PB1-PB2-HA constellations: VVY (B/Bankok/163/1990-like, 13 sequences isolated 1990 -- 5 Jan 1995), VYV (B/Nanchang/630/1994-like, 2 sequences isolated 1994 -- 1996) and VYY (B/New York/24/1993-like, 2 sequences isolated 8 Jan 1993 -- 1994).
An additional instance of reassortants with a YYV PB1-PB2-HA constellation (B/Waikato/6/2005-like, 16 sequences isolated 9 June -- 12 November in 2005) was discovered when investigating a larger dataset with only PB1, PB2, HA and NA sequences.

In all cases, PB1-PB2-HA reassortants have not persisted for prolonged periods of time and have not been fixed in the influenza B population.
In particular reassortment events combining PB1 and PB2 segments of different lineages, \textit{e.g.} B/Nanchang/630/1994-like and B/New York/24/1993-like isolates (each represented by two isolates), exhibit poor sampling and short circulation times.
In the case of B/Nanchang/630/1994-like viruses this is up to two years.
In contrast, reassortments combining PB1+2 and HA of different lineages (B/Bankok/163/1990-like and B/Waikato/6/2005-like isolates) have more isolates and circulated for much longer periods of time in the past: up to 4 years for B/Bankok/163/1990-like, although the more recently isolated B/Waikato/6/2005-like viruses circulated for only 5 months.

@@

Another possibility is the action of balancing selection in preserving the diversity in one segment, whilst the other segments hitchhike along.
A good candidate for this would be HA, as it is now the sole bearer of substantial antigenic diversity within the influenza B population, as the Vic lineage NA segment went extinct in 2002 (Figures \ref{tmrcaOT}, \ref{lineageRatiosOverTime} and \ref{railroadPlot})).
We find this scenario unlikely: if PB1+2 segments were hitchhiking with the HA segment stochastically we would expect to see more PB1+2 versus HA reassortants and fixation of Vic or Yam lineage PB1+2 segments in the influenza B population.
Our main dataset comprising 452 complete influenza B virus genomes has one instance of these kinds of reassortants characterised by B/Bankok/163/1990-like genome constellations.
The larger dataset with 1433 PB1, PB2 and HA sequences has an additional example of this: isolates with B/Waikato/6/2005-like PB1-PB2-HA constellations.
We note that in both cases the genome constellations of these reassortants were not fixed in the influenza B population and the more recent B/Waikato/6/2005-like reassortants persisted for a much shorter period of time (5 months versus 4 years), suggestive of declining PB1+2/HA reassortant fitness over time.

@@

\begin{figure}[h]
\caption{\textbf{Hierarchical clustering of amino acid sites on PB1 and PB2 segments based on $\chi^{2}_{df}$ values.}}
\label{ChiPB1PB2}
\end{figure}

\begin{figure}[h]
\caption{\textbf{Hierarchical clustering of amino acid sites on PB1 and HA segments based on $\chi^{2}_{df}$ values.}}
\label{ChiPB1HA}
\end{figure}

>>>>

\begin{figure}[h]
	\centering		
	\includegraphics[width=0.95\textwidth]{figures/RailroadPlotDated.pdf}
	\caption{\textbf{Schematic plot of reconstructed reassortments between Victoria and Yamagata lineage segments of influenza B virus.}
Lineages that coassort in genomes are represented by 8 parallel lines, with lineages that derive from the original Victoria clade colored red and lineages that derive from the original Yamagata clade colored blue.
Inter-lineage reassortment events are indicated by lines entering a different genome.
The angle of incoming lineages represents uncertainty in the timing of the event (mean date of the reassortant node and its parent node).
Lineage extinction dates are not shown accurately.}
	\label{railroadPlot}
\end{figure}

@@

We have identified 3 instances of mixed-lineage PB1-PB2-HA reassortants from the data with the following PB1-PB2-HA constellations: VVY (B/Bangkok/163/1990-like, 13 sequences isolated 1990 -- 5 Jan 1995), VYV (B/Nanchang/630/1994-like, 2 sequences isolated 1994 -- 1996) and VYY (B/New York/24/1993-like, 2 sequences isolated 8 Jan 1993 -- 1994).
An additional instance of reassortants with a YYV PB1-PB2-HA constellation (B/Waikato/6/2005-like, 16 sequences isolated 9 June -- 12 November in 2005) was discovered when investigating a larger dataset with only PB1, PB2, HA and NA sequences.

In all cases, PB1-PB2-HA reassortants have not persisted for prolonged periods of time and have not been fixed in the influenza B population.
In particular reassortment events combining PB1 and PB2 segments of different lineages, \textit{e.g.} B/Nanchang/630/1994-like and B/New York/24/1993-like isolates (each represented by two isolates), exhibit poor sampling and short circulation times.
In the case of B/Nanchang/630/1994-like viruses this is up to two years.
In contrast, reassortments combining PB1+2 and HA of different lineages (B/Bangkok/163/1990-like and B/Waikato/6/2005-like isolates) have more isolates and circulated for much longer periods of time in the past: up to 4 years for B/Bangkok/163/1990-like, although the more recently isolated B/Waikato/6/2005-like viruses circulated for only 5 months.

@@

Another possibility is the action of balancing selection in preserving the diversity in one segment, whilst the other segments hitchhike along.
A good candidate for this would be HA, as it is now the sole bearer of substantial antigenic diversity within the influenza B population, as the Vic lineage NA segment went extinct in 2002 (Figures \ref{tmrcaOT}, \ref{lineageRatiosOverTime} and \ref{railroadPlot})).
We find this scenario unlikely: if PB1+2 segments were hitchhiking with the HA segment stochastically we would expect to see more PB1+2 versus HA reassortants and fixation of Vic or Yam lineage PB1+2 segments in the influenza B population.
Our main dataset comprising 452 complete influenza B virus genomes has one instance of these kinds of reassortants characterised by B/Bangkok/163/1990-like genome constellations.
The larger dataset with 1433 PB1, PB2 and HA sequences has an additional example of this: isolates with B/Waikato/6/2005-like PB1-PB2-HA constellations.
We note that in both cases the genome constellations of these reassortants were not fixed in the influenza B population and the more recent B/Waikato/6/2005-like reassortants persisted for a much shorter period of time (5 months versus 4 years), suggestive of declining PB1+2/HA reassortant fitness over time.

@@

\section*{MCC trees}

\begin{figure}
\centering  
\includegraphics[width=0.95\textwidth]  {supp_figures/InfB_genomeGrid.pdf}
\caption{\textbf{MCC trees of 8 segments.}}
\label{}
\end{figure}

\section*{Linkage disequilibrium data}

\begin{figure}
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/Chi_PB1_PB2.png}
\caption{\textbf{Hierarchical clustering of amino acid sites on PB1 and PB2 segments based on $\chi^{2}_{df}$ values.}}
\label{ChiPB1PB2}
\end{figure}

\begin{figure}
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/Chi_PB1_HA.png}
\caption{\textbf{Hierarchical clustering of amino acid sites on PB1 and HA segments based on $\chi^{2}_{df}$ values.}}
\label{ChiPB1HA}
\end{figure}

\begin{figure}
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/Chi_PB2_HA.png}
\caption{\textbf{Hierarchical clustering of amino acid sites on PB2 and HA segments based on $\chi^{2}_{df}$ values.}}
\label{ChiPB2HA}
\end{figure}

We have identified amino acid sites which exhibit a high degree of linkage disequilibrium between PB1, PB2 and HA segments.
Sites with high LD on PB1, PB2 and HA segments could be interpreted as either sites which have drifted apart due to lack of reassortment between the three segments or as sites which prevent the cooperation of the three proteins at the molecular level and thus prevent inter-lineage PB1-PB2-HA reassortants from having comparable fitness to viruses with `pure' lineage PB1-PB2-HA complexes.

None of the sites we have identified on PB1 and PB2 proteins fall within the previously described regions that form the contacts between the two polymerase subunits \cite{sugiyama2009}.

Sites 211, 213 and 214 on the PB1 protein are very close to each other and the stretch of sequence around these residues contains many positively charged amino acids (lysine and arginine).
Multiple nuclear localization signals (NLSs) are predicted to occur around this region and sites 211, 213 and 214 are either predicted to be near the end of a mono-partite NLS or the beginning of a bi-partite NLS.
Previous research \cite{nath1990} suggests that in the influenza A PB1 protein residue 211 (homologous to influenza B PB1 residue 211) is the last residue of a bi-partite NLS.
Almost all Yamagata lineage isolates possess arginine (R) residue at PB1 positions 211 and 214 and a serine (S) residue at position 213, whereas Victoria lineage isolates have lysine (K) at positions 211 and 214 and threonine (T) at position 213.
It remains to be seen whether these sites significantly affect the nuclear import efficiency of the PB1 protein of either lineage.
Though the PB1 protein is known to accumulate in the nucleus on its own, efficient import into the nucleus requires the presence of the PA protein \cite{fodor2004}.
Similarly, sites 382 and 397 on the PB2 protein are close to residues 377, 406 and 408 which are homologous to sites in influenza A that are responsible for mRNA cap-binding \cite{guilligay2008}.
Again, we cannot determine whether this finding is significant or not.

\section*{Tree-to-tree comparisons}

Figures \ref{SPRdistancesTrees} and \ref{SPRdistancesReplicates} are showing approximate SPR distances prior to the normalization procedure (see Methods).
Figure \ref{SPRdistancesTrees} shows that MP and NS trees show greater approximate SPR distances when compared to any other segment tree, which is consistent with the fact that only M1 and NS1 coding sequences, which are shorter than any other coding sequences, were used to infer the trees.
Figure \ref{SPRdistancesReplicates} confirms this by showing greater variability in tree topology (as greater approximate SPR distance) when MP and NS trees are compared against their replicates.

For all other segments approximate SPR distances between all pairs of trees prior to normalization suggests that most segment pairs have similar posterior distributions of approximate SPR distances.

\begin{figure}
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_aSPRdistances_trees.png}
\caption{\textbf{Approximate SPR distances between all pairs of trees of segments.}}
\label{SPRdistancesTrees}
\end{figure}

\begin{figure}
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_aSPRdistances_replicates.png}
\caption{\textbf{Approximate SPR distances between replicate trees of each segment.}}
\label{SPRdistancesReplicates}
\end{figure}

Figures \ref{deltaTMRCAtrees} and \ref{deltaTMRCAreplicates} show $\Delta$TMRCA statistics for inter-segment and intra-segment, respectively, comparisons prior to normalization.
Figure \ref{deltaTMRCAtrees} shows that similarity in TMRCAs (as low $\Delta$TMRCA) between PB1, PB2 and HA segments is due to considerably smaller $\Delta$TMRCA statistic for the comparison between these segments.
It is also noteworthy that NA and MP comparisons have similarly low values for this statistic.
Figure \ref{deltaTMRCAreplicates} shows that TMRCAs are quite robust over the course of MCMC and node times do not, even in the most extreme cases, deviate from their replicates by more than 1 year.

\begin{figure}
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_deltaTMRCA_trees.png}
\caption{\textbf{Mean $\Delta$TMRCA between all pairs of trees of segments.}}
\label{deltaTMRCAtrees}
\end{figure}

\begin{figure}
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_deltaTMRCA_replicates.png}
\caption{\textbf{Mean $\Delta$TMRCA between replicate trees of each segment.}}
\label{deltaTMRCAreplicates}
\end{figure}

\begin{figure}
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_normRErate.png}
\caption{\textbf{Normalized reassortment rate (approximate number of SPR moves per total time in both trees) between all pairs of trees of segments.}}
\label{}
\end{figure}

\begin{figure}
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_RErate_trees.png}
\caption{\textbf{Reassortment rate (approximate number of SPR moves per total time in both trees) between all pairs of trees of segments.}}
\label{}
\end{figure}

\begin{figure}
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_RErate_replicates.png}
\caption{\textbf{Reassortment rate (approximate number of SPR moves per total time in the tree) between replicate trees of each segment.}}
\label{}
\end{figure}

\begin{figure}
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-PB2_hist.png}
\caption{\textbf{Difference}}
\label{}
\end{figure}

\begin{figure}
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-PB2_hist2.png}
\caption{\textbf{Ratio}}
\label{}
\end{figure}

\begin{figure}
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-HA_hist.png}
\caption{\textbf{Difference}}
\label{}
\end{figure}

\begin{figure}
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB1-HA_hist2.png}
\caption{\textbf{Ratio}}
\label{}
\end{figure}

\begin{figure}
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB2-HA_hist.png}
\caption{\textbf{Difference}}
\label{}
\end{figure}

\begin{figure}
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_supp_PB2-HA_hist2.png}
\caption{\textbf{Ratio}}
\label{}
\end{figure}

\begin{figure}
\centering  
\includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_PB1-PB2_corr.png}
\caption{\textbf{Correlation}}
\label{}
\end{figure}

\begin{figure}
\centering  
\includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_PB1-HA_corr.png}
\caption{\textbf{Correlation}}
\label{}
\end{figure}

\begin{figure}
\centering  
\includegraphics[width=0.5\textwidth]  {supp_figures/InfB_supp_PB2-HA_corr.png}
\caption{\textbf{Correlation}}
\label{}
\end{figure}

\clearpage

\begin{figure}
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_8x8_aaLD_MDS.png}
\caption{\textbf{2-dimensional MDS embedding of segment based on mean amino acid linkage disequilibrium ($\chi\_{df}^{2}$).}}
\label{}
\end{figure}

\begin{figure}
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_aaLD_MatrixMDScorr.png}
\caption{\textbf{Correlation between mean amino acid linkage disequilibrium and MDS embedding distance for all pairs of segments.}}
\label{}
\end{figure}

\begin{figure}
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_8x8_ntLD_MDS.png}
\caption{\textbf{2-dimensional MDS embedding of segment based on mean nucleotide linkage disequilibrium ($\chi\_{df}^{2}$)}}
\label{}
\end{figure}

\begin{figure}
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_ntLD_MatrixMDScorr.png}
\caption{\textbf{Correlation between mean nucleotide linkage disequilibrium and MDS embedding distance for all pairs of segments.}}
\label{}
\end{figure}
